---
title: "Housing Costs and Commuting Patterns Across the United States, 2021-2024"
author: Stanley Xiao
format: html
editor: visual
---

## Introduction

Housing costs and commuting are two important parts of everyday life that vary significantly across the United States. Differences in housing costs, income, geography, and commuting patterns result in very different experiences depending on which part of the country someone lives in.

In this project, I use data from the U.S. Census Bureau's Public Use Microdata Sample (PUMS) API to investigate housing costs and commuting patterns across the United States from 2021 through 2024. I'm particularly interested in examining how gross rent, rent as a percentage of household income, and commute duration vary across geographic areas and over time.

I created a set of reusable R functions to obtain and process data from the Census API. I also developed custom summary and plotting methods that account for the person weights included in the dataset. I used these functions to investigate patterns in housing costs and commuting across different geographic areas and years.

## Data Source

The data for this project comes from the U.S. Census Bureau's Public Use Microdata Sample (PUMS) API. PUMS provides person-level-like records containing demographic, housing, education, and commuting information. This project focuses on data from 2021, 2022, 2023, and 2024.

## Setup

```{r}
#Install packages as necessary
library(httr)
library(jsonlite)
library(lubridate)
library(tidyverse)
```

## Processing Census Variable Information

The variables returned by the Census API are usually in the form of numeric codes rather than descriptive labels. The first step is to obtain the variable metadata for each year which will then be converted into meaningful values later in the analysis.

The state variable also differs across survey years. The 2021 and 2022 data use 'ST' while the 2023 and 2024 data use 'STATE'. The following function accounts for this difference when retrieving the required variable information.

```{r}
variables_needed <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "REGION", "DIVISION"
)

get_variable_info <- function(year) {
  #Check that the year is valid
  if (!year %in% 2021:2024) {
    stop("Year must be 2021/2022/2023/2024")
  }
  
  #URL for selected year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )
  
  #Query Census API
  response <- GET(url)
  
  #Parse JSON
  variable_data <- content(
    response,
    as = "parsed",
    type = "application/json"
  )
  
  #State variable differs depending on year
  state_var <- if (year <= 2022) {
  "ST"
  } else {
    "STATE"
  } 
  
  #Add the appropriate state variable
  vars <- c(variables_needed, state_var)
  
  #Keep only variables needed for the project
  variable_data$variables[vars]
}

#Variable information for each year
variables_2021 <- get_variable_info(2021)
variables_2022 <- get_variable_info(2022)
variables_2023 <- get_variable_info(2023)
variables_2024 <- get_variable_info(2024)

#Combine yearly variable info into one list
census_variables <- list(
  '2021' = variables_2021,
  '2022' = variables_2022,
  '2023' = variables_2023,
  '2024' = variables_2024
)

#Check for categorical variable value (FER, HHL, SCH, SCHL, SEX)
variables_2024$SEX$values$item
```

```{r}
#Create data folder
dir.create("data", showWarnings = FALSE)

#Save variable information
saveRDS(census_variables, "data/census_variables.rds")
```

The saved metadata (`census_variables`) will serve as a reference for interpreting the Census API data, allowing categorical variables to be converted into descriptive factor levels.

##Querying Single-Year Census Data

Next, we implement functions to query single-year PUMS data and convert the response into a tibble. The function allow users to select between numeric, categorical, and geographic variables.

#### First helper API response$\to$tibble
```{r}
response_to_tibble <- function(response) {
  
  #Parse JSON response
  census_data <- content(
    response,
    as = "text",
    encoding = "UTF-8"
  ) |>
    fromJSON()
  
  #First row contains variable names
  variable_names <- census_data[1,]
  
  #Remaining rows are observation values
  census_data <- census_data[-1, , drop = FALSE]
  
  #Add variable names
  colnames(census_data) <- variable_names
  
  #Convert to tibble
  census_data <- as_tibble(census_data)
  names(census_data) <- variable_names
  
  census_data
}
```

#### Second helper Census time interval$\to$midpoint
```{r}
#Convert Census time interval to midpoint
time_midpoint <- function(time_interval) {
  
  #Return NA for missing values
  if (is.na(time_interval) || length(time_interval) == 0) {
    return(NA)
  }
  
  #Split beginning and ending times
  times <- strsplit(time_interval, " to ")[[1]]
  
  #Remove periods from a.m and p.m.
  times <- gsub("\\.", "", times)
  
  #Convert to times
  start_time <- parse_date_time(times[1], "I:M p")
  end_time <- parse_date_time(times[2], "I:M p")
  
  #Calculate midpoint
  midpoint <- start_time + ((end_time - start_time) / 2)
  
  return(midpoint)
}

#Check what is outputted if there are any obvious NA values we can check. Results did not show any NA values.
#head(variables_2024$JWDP$values$item)
#tail(variables_2024$JWDP$values$item)
#head(variables_2024$JWAP$values$item)
#tail(variables_2024$JWAP$values$item)
```

#### Third helper to retrieve census data
```{r}
#API Key is now required to access Census data
census_api_key <- Sys.getenv("CENSUS_API_KEY")
get_census_data <- function(
    year = 2024,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geo_value = NULL,
    key = census_api_key
) {
  
  #Allowed variables
  allowed_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  
  allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  
  allowed_geography <- c("Region", "Division", "State")
  
  #Check year
  if (length(year) != 1 || !year %in% 2021:2024) {
    stop("Year must be 2021/2022/2023/2024")
  }
  
  #Check numeric variables
  if (!all(numeric_vars %in% allowed_numeric)) {
    stop("Invalid numeric variable")
  }
  
  if (!"PWGTP" %in% numeric_vars) {
    stop("PWGTP must be included as a numeric variable")
  }

  if (length(setdiff(numeric_vars, "PWGTP"))<1) {
    stop("At least one numeric variable besides PWGTP must be selected")
  }

  #Check categorical variables
  if (length(categorical_vars) < 1 || !all(categorical_vars %in% allowed_categorical)) {
    stop("At least one valid categorical variable must be selected")
  }
  #Check geography
  if (length(geography) != 1 || !geography %in% allowed_geography) {
    stop("Geography must be Region, Division, or State")
  }
  
  #Read variable metadata
  variable_info <- readRDS("data/census_variables.rds")[[as.character(year)]]
  
  #Determine Census geography variable
  geo_var <- if (geography == "State") {
    if (year <= 2022) "ST" else "STATE"
  } else {
    toupper(geography)
  }
  
  #Obtain allowable geography values
  geo_values <- variable_info[[geo_var]]$values$item
  
  #Use first available geography as default
  if (is.null(geo_value)) {
    geo_value <- names(geo_values)[1]
  }
  
  #Check selected geography value
  if (tolower(geo_value) != "all" &&
      !geo_value %in% names(geo_values)) {
    stop("Invalid geography value selected")
  }
  
  #Variables requested from API
  selected_vars <- unique(c(numeric_vars, categorical_vars))
  
  #Create Census API URL
  url <- paste0("https://api.census.gov/data/",
  year,
  "/acs/acs1/pums"
  )
  
  #Allow one geography value or all values
  if(tolower(geo_value) == "all") {
    geography_query <- "*"
  } else {
    geography_query <- geo_value
  }
  
  #Query Census API
  response <- GET(url, query = list(
    get = paste(selected_vars, collapse = ","),
    'for' = paste0(tolower(geography), ":", geography_query),
    key = census_api_key
  ))
  
  #Check API request as part of debugging process
  #print(response$url) #API key will be included in this line
  
  #Convert response to tibble
  census_data <- response_to_tibble(response)
  
  #Convert numeric variables (JWAP & JWDP are time variables and not ordinary variables)
  regular_numeric <- intersect(
    numeric_vars,
    c("AGEP", "GASP", "GRPIP", "JWMNP", "PWGTP")
  )
  
  census_data[regular_numeric] <- lapply(
    census_data[regular_numeric],
    as.numeric
  )
  
  #Convert categorical variables to factors
  for (variable in categorical_vars) {
    
    value_labels <- variable_info[[variable]]$values$item
    
    census_data[[variable]] <- factor(
      census_data[[variable]],
      levels = names(value_labels),
      labels = unlist(value_labels)
    )
  }
  
  ##Convert time variables to midpoint
  time_vars <- intersect(
    numeric_vars,
    c("JWAP", "JWDP")
  )
  
  for (variable in time_vars) {
    
    #Get time interval labels 
    time_labels <- variable_info[[variable]]$values$item
    
    #Match Census codes to time intervals
    intervals <- sapply(
      census_data[[variable]],
      function(code) {
        if (is.na(code) || !code %in% names(time_labels)) {
          NA_character_
        } else {
          time_labels[[code]]
        }
      }
    )
    
    #Convert intervals to midpoint
    census_data[[variable]] <- lapply(
      intervals,
      time_midpoint
    ) |>
      unlist()
  }
  
  #Add census class
  class(census_data) <- c("census", class(census_data))

  return(census_data)
  #End of get_census_data function
}

#Test cases used during development
test_census <- get_census_data()
#head(test_census)
class(test_census)
```

## Combining Multiple Years

The single-year function can be extended to compare Census data across multiple survey years. The following function allows the user to select multiple years and combines the results into one tibble.

```{r}
get_census_years <- function(
    years = 2021:2024,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geo_value = NULL
) {
  #Allow user to select all years
  if (length(years) == 1 && tolower(as.character(years)) == "all") {
    years <- 2021:2024
  }
  
  #Check years
  if (!all(years %in% 2021:2024)) {
    stop("years must be 2021/2022/2023/2024 or all")
  }
  
  #Empty list to store each year
  census_list <- list()
  
  #Get data from each year
  for (year in years) {
    
    census_list[[as.character(year)]] <- get_census_data(
      year = year,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography = geography,
      geo_value = geo_value
    )
    
    #Add year variable
    census_list[[as.character(year)]]$year <- factor(year)
  }
  
  #Combine years into one tibble
  census_data <- bind_rows(census_list)
  
  #Add census class
  class(census_data) <- c("census", class(census_data))
  
  return(census_data)
}

#Test cases used during development
test_multiple <- get_census_years( years = c(2022, 2024))
#head(test_multiple)
class(test_multiple)
#table(test_multiple$year)
test_all_years <- get_census_years(years = "all")
class(test_all_years$year)
```

## Generic Census Summary Function

Because each row of the PUMS data may represent multiple people, the PWGTP variable must be considered when summarizing the data. The below custom summary.census() method calculates weighted means and standard deviations for numeric variables and weighted counts for categorical variables.

```{r}
summary.census <- function(
    object,
    numeric_vars = NULL,
    categorical_vars = NULL
) {
  #Find numeric variables if none are specified
  if (is.null(numeric_vars)) {
    numeric_vars <- names(object)[sapply(object, is.numeric)]
    
    #Don't summarize weight or time variables
    numeric_vars <- setdiff(
      numeric_vars,
      c("PWGTP", "JWAP", "JWDP")
    )
  }
  #Find categorical variables if none are specified
  if (is.null(categorical_vars)) {
    categorical_vars <- names(object)[sapply(object, is.factor)]
  }
  #Store results
  results <- list()
  
  #Summarize numeric variables
  for (variable in numeric_vars) {
    x <- object[[variable]]
    weights <- object$PWGTP
    
    #Remove missing values
    keep <- !is.na(x) & !is.na(weights)
    x <- x[keep]
    weights <- weights[keep]
    
    #Weighted mean
    weighted_mean <- sum(x * weights) / sum(weights)
    
    #Weighted standard deviation
    weighted_sd <- sqrt(
      (sum((x^2) * weights) / sum(weights)) - (weighted_mean^2)
    )
    results[[variable]] <- c(mean = weighted_mean, sd = weighted_sd)
  }
  #Summarize categorical variables
for (variable in categorical_vars) {
  counts <- tapply(
    object$PWGTP,
    object[[variable]],
    sum,
    na.rm = TRUE
  )
  results[[variable]] <- counts
  }

  return(results)
}

#Test case
summary(test_all_years)
summary(test_all_years, numeric_vars = "AGEP", categorical_vars = "year")
```
## Generic Census Plot Function
A custom plot.census() method can also be used to compare a numeric variable across the levels of a categorical variable. The person weights are included in the plot so that each observation is represented according to its PUMS weight.

```{r}
plot.census <- function(
    x,
    cat_var,
    num_var
) {
  #Check categorical variable
  if (!cat_var %in% names(x)) {
    stop("Categorical variable cannot be found")
  }
  
  #Check numeric variable
  if (!num_var %in% names(x)) {
    stop("Numeric variable cannot be found")
  }
  
  #Create weighted boxplot
  ggplot(
    x,
    aes(
      x = get(cat_var),
      y = get(num_var),
      weight = PWGTP
    ) 
  ) +
    geom_boxplot() +
    labs(
      x = cat_var,
      y = num_var
    )
}

#Test case
plot(
  test_all_years,
  cat_var = "year",
  num_var = "AGEP"
)

plot(
  test_all_years,
  cat_var = "SEX",
  num_var = "AGEP"
)
```

## Investigation

For the final investigation, I use the functions developed above to examine housing costs and commuting patterns across the United States from 2021 through 2024. I focus on differences across Census divisions and over time.

The primary variables used in this investigation are gross rent (`GASP`), gross rent as a percentage of household income (`GRPIP`), and travel time to work (`JWMNP`). I also include sex (`SEX`) as a categorical variable and `PWGTP` so that the observations can be properly weighted.

```{r}
#Retrieve data for all Census divisions from 2021-2024
investigation_data <- get_census_years(
  years = 2021:2024,
  numeric_vars = c("GASP", "GRPIP", "JWMNP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "Division",
  geo_value = "all"
)

#The above threw an error, let's go back and see what went wrong and break it down
test_investigation <- get_census_years(
  years = 2024,
  numeric_vars = c("GASP", "GRPIP", "JWMNP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "Division",
  geo_value = 1
)

exists("test_investigation")
dim(test_investigation)
head(investgation_data)
names(investgation_data)
dim(investgation_data)
```

```{r}
#Saving locally since pulling via API takes a while
saveRDS(investigation_data, "data/investigation_data.rds")
```

