---
title: "Housing Costs and Commuting Patterns Across the United States, 2021-2024"
author: Stanley Xiao
format: html
editor: visual
---

## Introduction

Housing costs and commuting are two important parts of everyday life that vary significantly across the United States. Differences in housing costs, income, geography, and commuting patterns result in very different experiences depending on which part of the country someone lives in.

In this project, I use data from the U.S. Census Bureau's Public Use Microdata Sample (PUMS) API to investigate housing costs and commuting patterns across the United States from 2021 through 2024. I'm particularly interested in examining how gross rent, rent as a percentage of household income, and commute duration vary across geographic areas and over time.

I created a set of reusable R functions to obtain and process data from the Census API. I also developed custom summary and plotting methods that account for the person weights included in the dataset. I used these functions to investigate patterns in housing costs and commuting across different geographic areas and years.

## Data Source

The data for this project comes from the U.S. Census Bureau's Public Use Microdata Sample (PUMS) API. PUMS provides person-level-like records containing demographic, housing, education, and commuting information. This project focuses on data from 2021, 2022, 2023, and 2024.

## Setup

```{r}
#Install packages as necessary
library(httr)
library(jsonlite)
library(lubridate)
library(tidyverse)
```

## Processing Census Variable Information

The variables returned by the Census API are usually in the form of numeric codes rather than descriptive labels. The first step is to obtain the variable metadata for each year which will then be converted into meaningful values later in the analysis.

The state variable also differs across survey years. The 2021 and 2022 data use 'ST' while the 2023 and 2024 data use 'STATE'. The following function accounts for this difference when retrieving the required variable information.

```{r}
variables_needed <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "REGION", "DIVISION"
)

get_variable_info <- function(year) {
  #Check that the year is valid
  if (!year %in% 2021:2024) {
    stop("Year must be 2021/2022/2023/2024")
  }
  
  #URL for selected year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )
  
  #Query Census API
  response <- GET(url)
  stop_for_status(response)
  
  #Parse JSON
  variable_data <- content(
    response,
    as = "parsed",
    type = "application/json"
  )
  
  #State variable differs depending on year
  state_var <- if (year <= 2022) {
  "ST"
  } else {
    "STATE"
  } 
  
  #Add the appropriate state variable
  vars <- c(variables_needed, state_var)
  
  #Keep only variables needed for the project
  variable_data$variables[vars]
}

#Variable information for each year
variables_2021 <- get_variable_info(2021)
variables_2022 <- get_variable_info(2022)
variables_2023 <- get_variable_info(2023)
variables_2024 <- get_variable_info(2024)

#Combine yearly variable info into one list
census_variables <- list(
  '2021' = variables_2021,
  '2022' = variables_2022,
  '2023' = variables_2023,
  '2024' = variables_2024
)

#Check for categorical variable value
variables_2024$SEX$values$item
```

```{r}
#Create data folder
dir.create("data", showWarnings = FALSE)

#Save variable information
saveRDS(census_variables, "data/census_variables.rds")
```

The saved metadata (`census_variables`) will serve as a reference for interpreting the Census API data, allowing categorical variables to be converted into descriptive factor levels.
